---
title: "JHDS_RR_PA1"
author: "Shean Dalton"
date: "Sunday, March 15, 2015"
output: html_document
  
---
INTRO:
  Learn about value of activity monitoring devices. Why would I want one, and I don't right now.  
  
  **Prepare workarea environmnet**  
  
 ```{r, setup, echo=TRUE}

  library(knitr)
  options(warn=-1)

```

 ```{r, alllibraries, echo=TRUE}

  if (!require("data.table")) {
      install.packages("data.table", repos="http://cran.rstudio.com/")
     library(data.table)
  }
 

  if (!require("sqldf")) {
    install.packages("sqldf", repos="http://cran.rstudio.com/")
    library(sqldf)  
  }
  
  
  if (!require("dplyr")) {
    install.packages("dplyr", repos="http://cran.rstudio.com/")
    library(dplyr)  
  }

  if (!require("xtable")) {
    install.packages("xtable", repos="http://cran.rstudio.com/")
    library(xtable)
  }
 
  if (!require("ggplot2")) {
    install.packages("ggplot2", repos="http://cran.rstudio.com/")
    library(ggplot2)
  }

  if (!require("lubridate")) {
    install.packages("lubridate", repos="http://cran.rstudio.com/")
    library(lubridate)  
  }  

  if (!require("sqldf")) {
    install.packages("sqldf", repos="http://cran.rstudio.com/")
    library(sqldf) 
  }  
  
  if (!require("lattice")) {
    install.packages("lattice", repos="http://cran.rstudio.com/")
    library(lattice) 
  }  

  require(lattice)
  require(sqldf)
  require(lubridate)
  require(ggplot2)
  require(xtable)
  require(dplyr)
  require(sqldf)
  require(data.table)

```

  **Loading and preprocessing the data**  
  1. LOAD the data Extract data from Internet  
 ```{r, echo=TRUE}
  setwd("C:/Users/Shean/Documents/GitHub/JHDS_RR_PA1/JHDS_RR_PA1")
  temp <- tempfile()
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  file <- "activity.csv"

  setInternet2(use = TRUE)

  download.file(url, temp)
  unzip(temp, files = NULL, list = FALSE, overwrite = TRUE,
          junkpaths = FALSE, exdir = ".", unzip = "internal",
          setTimes = FALSE)
  
  if (file.exists(file)==FALSE){
    temp <- tempfile()
    download.file(url, temp)
    # unzip(temp, list=TRUE)
    unzip(temp, files = NULL, list = FALSE, overwrite = TRUE,
          junkpaths = FALSE, exdir = ".", unzip = "internal",
          setTimes = FALSE)
  }
  unlink(temp)

  rm(url)
  rm(file)
  rm(temp)
  
  data <- read.csv("activity.csv", header = TRUE, na.strings = "na", stringsAsFactors=FALSE)
  data <- data.table(data)
 ```

  2. PROCESS/Transform data: Add date_time column, add Steps interger column** 
```{r, echo=TRUE}
  data$date_time <- sprintf("%04d", data$interval)
  data$date_time <- as.character(sub("(?<=.{2})", ":", data$date_time, perl=TRUE))
  data$date_time <- as.character(sub("(?<=.{5})", ":00", data$date_time, perl=TRUE))
  data$date_time <- paste(data$date, data$date_time, sep = " ")  
  data$date_time <- ymd_hms(data$date_time)
  data$steps_int <- as.numeric(data$steps)  
````  
  
  - Transform data: Populate new table "data_no_NA" with no NA rows** 
```{r, echo=TRUE}
  row_has_na <- apply(data, 1, function(x){any(is.na(x))})
  data_no_NA <- data[!row_has_na,]
  data_no_NA$date <- as.Date(data_no_NA$date)
```  
  
  - Transform data: Populate new table "steps_day", aggregates date and sums steps_in**
```{r, echo=TRUE}
  
  steps_day <- aggregate(data_no_NA$steps_int, by = list(data_no_NA$date), FUN=sum, na.rm=TRUE)
  steps_day <- data.table(steps_day)
  setnames(steps_day, "Group.1", "Date")
  steps_day$Date <- as.Date(steps_day$Date)
  setnames(steps_day, "x", "Steps")
  steps_day$Steps <- as.numeric(steps_day$Steps)

```

  **What is mean total number of steps taken per day?**

  1. TOTAL number of steps taken per day
  
```{r, ggplot2_total_number_of_steps_taken_per_day, echo=TRUE, fig.path='figure/'}
  
  plot1 <- (steps_day)
  g <-   ggplot(steps_day, aes(x=Date, y=Steps)) + geom_bar(stat="identity") + 
    labs(x="Day", y="Steps")
  print(g)
```

  2. HISTOGRAM of the total number of steps taken per day

```{r, Histogram_total_number_of_steps_taken_per_day, echo=TRUE, fig.path='figure/'}
  plot1 <- (steps_day$Steps) 
  hist(plot1, breaks=12, main="Steps Each Day",xlab="Steps", col="red")
  rug(plot1,side=1)
```

  3. MEAN and Median of the total number of steps taken per day  
  
  - Mean  
```{r, echo = TRUE}
  print(as.integer(mean(steps_day$Steps))+0.5)
````

  - Median  
```{r, echo = TRUE}
    print(as.integer(median(steps_day$Steps))+0.5)
```

  **What is the average daily activity pattern?**  

  - Transform data: : Populate new table "steps_interval_mean", aggregates intervals and sums steps_in   

```{r, echo = TRUE}  
  steps_interval_mean <- aggregate(data_no_NA$steps_int, by = list(data_no_NA$interval), FUN=mean, na.rm=TRUE)
  steps_interval_mean <- data.table(steps_interval_mean)
  setnames(steps_interval_mean, "Group.1", "Interval")
  setnames(steps_interval_mean, "x", "Steps_mean")
```

  1. TIME series plot: Average Steps at Each Interval

```{r, plot_average_steps_at_each_interval, echo = TRUE, fig.path='figure/'}
  x <- steps_interval_mean$Interval
  y <- steps_interval_mean$Steps_mean
  p <- plot(x, y, main = "Average Steps At Each Interval", type = "l", xlab = "(5-min. Intervals)", ylab = "(Average Steps)")
  print(p)
```

  2. INTERVAL on average with Max Steps, and Max Steps Count: measured from across days monitored  

```{r, echo = TRUE}
    interval_max_mean <- filter(steps_interval_mean, steps_interval_mean$Steps_mean == max(steps_interval_mean$Steps_mean))
    interval_max_mean <- data.table(interval_max_mean)
    print(interval_max_mean)
```

  **Imputing Missing Values**  
  
  - Transform Data: populate table "row_with_NA" that includes only rows that include NA values  

```{r, echo = TRUE}
  row_with_NA <- data[!row_has_na,]
```

  1. TOTAL Number of Data Entries with Missing Values  

```{r, echo = TRUE}
  print(nrow(row_with_NA))
```

  2. STRATEGY Convert NA Step values to their respective interval mean (integers)  

  3. CREATE a new data set  
  
  - Transform: populate table "idata" that includes "data" table and "respective interval means"  

```{r, echo = TRUE}
  
  setnames(steps_interval_mean, "Interval", "interval")

  idata <- merge(x = data, y = steps_interval_mean, by = "interval", all.x=TRUE) 
  
  idata$Steps_mean <- as.integer(idata$Steps_mean+0.5)

  idata <- 
    idata %>%
            mutate(steps=ifelse(is.na(steps)|steps == "NA", Steps_mean, steps))
 
  idata <- idata %>% select(interval, steps, date, date_time, Steps_mean)

  idata$steps <- as.integer(idata$steps)

```    
    - Transform data: Populate new table "isteps_day", aggregates date and sums steps  
     
```{r, echo = TRUE}
  isteps_day <- aggregate(idata$steps, by = list(idata$date), FUN=sum, na.rm=TRUE)
  isteps_day <- data.table(isteps_day)
  setnames(isteps_day, "Group.1", "Date")
  setnames(isteps_day, "x", "Steps")
```

  4. HISTOGRAM of the total number of steps taken each day

```{r, Histogram_total_number_of_steps_taken_per_day_imputed_data, echo = TRUE, fig.path='figure/'}
  x <- (isteps_day$Steps)
  hist(x, breaks=12, main="Imputed Steps Each Day",xlab="Imputed Steps", col="red")
  rug(x,side=1)
```  
  
  - Calculate and report the mean steps per day of imputed data  

```{r, echo = TRUE}
  print(mean(isteps_day$Steps))
```  
  
  - Calculate and report the median steps per day of imputed data  

```{r, echo = TRUE}
  print(median(isteps_day$Steps))
```  
  
  - Do these values differ from the estimates from the non-imputed data?  
    
```{r, plot_comparing_imputed_to_nonimputed_data, echo = TRUE, fig.path='figure/'}
  x <- isteps_day$Date
  y <- isteps_day$Steps
  xi <- steps_day$Date
  yi <- steps_day$Steps
  plot(xi, yi, type="n", xlab="Date", ylab="Steps", cex.lab=".75", main = "Imputed Overlay to Non Imputed Steps Per Interval")
  points(x, y, type="l", col="red")
  points(xi, yi, type="l", col="blue")
  legend("topleft",legend = c("Imputed Steps", "Non Imputed Steps"), col = c("red", "blue"), lwd = .75, cex = .65, bty="n")
```  
  
  - Impact of imputing missign data on the estimates of the total daily number of steps  
  
  - Mean for interval did not change: comparing interval non-imputed means to imputed means  
  
  - Transform data: Populate new table "isteps_interval_mean", aggregates intervals and sums steps

```{r, echo = TRUE}
  isteps_interval_mean <- aggregate(idata$steps, by = list(idata$interval), FUN=mean, na.rm=TRUE)
  isteps_interval_mean <- data.table(isteps_interval_mean)
  setnames(isteps_interval_mean, "Group.1", "Interval")
  setnames(isteps_interval_mean, "x", "Steps_mean")
  i_interval_max_mean <- filter(isteps_interval_mean, isteps_interval_mean$Steps_mean == max(isteps_interval_mean$Steps_mean))
  i_interval_max_mean <- data.table(i_interval_max_mean)
```  
  
  - Imputed MAX mean for intervals and days 
```{r, echo = TRUE}
  print(i_interval_max_mean)

```  
  
  
  - NON-Imputed MAX mean for intervals and days 
  
```{r, echo = TRUE}
  print(interval_max_mean)
``` 
  
  - Mean for days did change: comparing interval non-imputed means to imputed means   
  - Imputed mean for days  
      
```{r, echo = TRUE}
  print(mean(isteps_day$Steps))
``` 
  
  
  - NON-Imputed mean for days  
      
```{r, echo = TRUE}
  print(mean(steps_day$Steps))
```  
  
  - Transform data: Populate new table "iDiff", aggregates intervals and sums steps  
      


  **Differences in Activity Patterns Between Weekdays and Weekends**  
  
  1. Transform data: add column "wend" to imputed data table "idata"  
    
```{r, echo = TRUE}
  idata$wend <- as.factor(ifelse(weekdays( idata$date_time) %in% c("Saturday","Sunday"), "Weekend", "Weekday"))
```  
  
  - Transform data: populate table "isteps_interval_mean_wk$Steps_mean"  
    
```{r, echo = TRUE}
  isteps_interval_mean_wk <- aggregate(idata$steps, by = list(idata$interval, idata$wend), FUN=mean, na.rm=TRUE)
  setnames(isteps_interval_mean_wk, "Group.1", "Interval")
  setnames(isteps_interval_mean_wk, "x", "Steps_mean")
  setnames(isteps_interval_mean_wk, "Group.2", "Weekday")  
  isteps_interval_mean_wk$Steps_mean <- as.integer(isteps_interval_mean_wk$Steps_mean+0.5)
```
 2. Panel plot of the 5-minute interval and average number of steps taken, averaged across all weekday days or weekend days  

<<<<<<< HEAD
```{r, panelplot_weekend_weekday, echo = TRUE, fig.path='figure/'}
=======
```{r, latticeistepsintervalmeanwk, echo = TRUE, fig.height=5}
>>>>>>> 9f1801c035ebcdf2e956a55e8947c6089b1589ba
  g <- xyplot(Steps_mean ~ Interval | Weekday, data = isteps_interval_mean_wk, type = "l" , layout = c(1,2), xlab="Time of day", ylab="Average steps", 
              scales=list( Steps_mean=list(tick.number=5)),
              main="Average steps taken per day"
           )
  print(g)

```